---
title: "PML - Course Project"
author: "Yigit Ozan Berk"
date: "9/22/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

# Initiation

```{r}
dat = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")

dat.test = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```



The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


# PreProcessing

```{r}
library(dplyr)
colMeans(is.na(dat))
```
There is a significant number of systematic NAs

```{r}
colMeans(is.na(dat))[colMeans(is.na(dat))> 0]
```

So 67 of columns are basically NA columns, those features are dropped.

```{r}
nacol = colMeans(is.na(dat))> 0.9
dat2 = dat[, !nacol]
#also drop user name, and time variables
dat2 = dat2[, -c(2, 3, 4, 5)]
#what is new_window and num_window column representing? new_window has the same number of "no"s as the number of NAs in the omitted NA columns.

#removing near zero variance columns
dat3 = dat2[, -nearZeroVar(dat2)]
#we are left with 55 columns
colnames(dat3)
```

Parallel Processing initiation

Parallel processing in caret can be accomplished with the parallel and doParallel packages. The following code loads the required libraries (note, these libraries also depend on the iterators and foreach libraries).
```{r}
library(parallel)
library(doParallel)
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```

The most critical arguments for the trainControl function are the resampling method method, the number that specifies the quantity of folds for k-fold cross-validation, and allowParallel which tells caret to use the cluster that we've registered in the previous step.
```{r}
fitControl <- trainControl(method = "cv",
                           number = 5,
                           allowParallel = TRUE)
```

Next, we use caret::train() to train the model, using the trainControl() object that we just created.
```{r}
set.seed(1)
rfFit <- train(classe ~ ., method="rf",data=dat3,trControl = fitControl)
```

After processing the data, we explicitly shut down the cluster by calling the stopCluster() and registerDoSEQ() functions. The registerDoSEQ() function is required to force R to return to single threaded processing.
```{r}
stopCluster(cluster)
registerDoSEQ()
```

```{r}
rfFit
```

testing the test set. what is the problem ID column?
```{r}
names = colnames(dat3)
intest1 = colnames(dat.test) %in% names
test1 = dat.test[, intest1]
```

